EBASE manuscript

### 2023-04-18 manu-draft.qmd
edit to final ebase v odum paragraphs

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange. Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.

<!-- edit -->

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding production and respiration with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods. Half of the slope estimates were significantly different for estimates from both the observed and detided dissolved oxygen, with production and gas exchange differing for the observed input and net metabolism and gas exchange differing for the detided input.  The RMSE of each regression comparison were generally lower in the detided results for all comparisons. This latter comparison suggests that detiding the time series produces more comparable estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]). Additionally, the effects of tidal advection produced a wider range of estimates for NEM in both methods, whereas the detided results were mostly negative indicating that conclusions about net autotrophy or heterotrophy may be inaccurate using observed dissolved oxygen as input. Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

<!-- edit -->

>>>>

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach from @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, with the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differences in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange. Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates using the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding production with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods. The slope estimates for respiration and gas exchange were significantly different for estimates from the observed dissolved oxygen, where all but respiration were signicantly different using the detided dissolved oxygen. The RMSE values of each regression comparison were generally lower in the detided results for all comparisons, excluding gas exchange which was higher with the detided dissolved oxygen compared to the observed. This latter comparison suggests that detiding the time series produces more comparable estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]). Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.
